---
title: "<b>Week 7: Example - comparing coding schemes </b>"
subtitle: "RMS2<br><br> "
author: ""
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
```


# Topics for today

+ Last time we spoke about the theory behind different contrasts.

+ Now we will look practically at the implementation and differences

---
# The data
```{r}
hosp_tbl <- read_csv("hospital.csv", col_types = "dff")
hosp_tbl %>%
  select(SWB, Treatment) %>%
  slice(1:10)
```

---
# Group Means

```{r}
hosp_tbl %>%
  select(1:2) %>%
  group_by(Treatment) %>%
  summarise(
    mean = round(mean(SWB),3),
    sd = round(sd(SWB),1),
    N = n()
  )
```

---
# Dummy (reference) model

```{r}
summary(lm(SWB ~ Treatment, data = hosp_tbl))
```

---
# Dummy (reference) model

.pull-left[

```{r, echo=FALSE}
dres <- summary(lm(SWB ~ Treatment, data = hosp_tbl))
round(dres$coefficients[,1],3)
```

+ Recall the equations for the group means:

$$\mu_{treatmentA} = \beta_0$$
$$\mu_{treatmentB} = \beta_0 + \beta_1$$ 

$$\mu_{treatmentC} = \beta_0 + \beta_2$$

]

.pull-right[

```{r, echo=FALSE}
hosp_tbl %>%
  select(1:2) %>%
  group_by(Treatment) %>%
  summarise(
    mean = round(mean(SWB),3)
  ) %>%
  kable(.) %>%
  kable_styling(., full_width = F)

```

]


---
# Effects (sum to zero) model

+ We need to change the contrast scheme from default.

```{r}
contrasts(hosp_tbl$Treatment) <- contr.sum 
contrasts(hosp_tbl$Treatment)
```

---
# Effects (sum to zero) model
```{r}
summary(lm(SWB ~ Treatment, data = hosp_tbl))
```

---
# Effects (sum to zero) model

.pull-left[
```{r, echo=FALSE}
Eres <- summary(lm(SWB ~ Treatment, data = hosp_tbl))
round(Eres$coefficients[,1],3)
```

+ Recall the equations for the group means:


$$\beta_0 = \frac{\mu_1 + \mu_2 + \mu_3}{3}$$

$$\mu_1 = \beta_0 + \beta_1$$

$$\mu_2 = \beta_0 + \beta_2$$

$$\mu_3 = \beta_0 - (\beta_1 + \beta_2)$$
]


.pull-right[

```{r, echo=FALSE}
hosp_tbl %>%
  select(1:2) %>%
  group_by(Treatment) %>%
  summarise(
    mean = round(mean(SWB),3)
  ) %>%
  mutate(
    Gmean = round(mean(mean),3)
  ) %>%
  kable(.) %>%
  kable_styling(., full_width = F)
```


]

---
# Manual contrast testing
+ We can structure a wide variety of contrasts so long as they can be written

1. A as a linear combination of population means
2. The associated coefficients sum to zero

+ So

$$H_0: c_1\mu_1 + c_1\mu_2 + c_3\mu_3 $$

+ With

$$c_1 + c_2 + c_3 = 0$$

---
# Manual contrast testing
+ For example, say we wanted to compare TreatmentA to TreatmentB and TreatmentC combined.

+ Our null becomes:

$$H_0: \mu_{treatmentA} = \frac{1}{2}(\mu_{treatmentB} + \mu_{treatmentC})$$

+ Or, is the average SWB of TreatmentA group equal to the average SWB of the combined TreatmentB and TreatmentC

+ We can set the contrasts here:
  + $c_1 = 1$
  + $c_2 = -1/2$
  + $c_3 = -1/2$

---
# Summary of today

+ We have looked at how to analyse experimental designs using different coding and contrasts.

+ Key take home:
  + Our $\beta$'s will represent differences in group means.
  + The scheme we use determines which groups.
  + This makes it very flexible for testing hypotheses.