---
title: "Principal Components Analysis"
subtitle: "Research Methods and Statistics 2<br><br> "
author: "ALEXANDER WEISS"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: TRUE
      countIncrementalSlides: FALSE

---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
library(MASS)
library(psych)
library(ggplot2)
library(gridExtra)
library(psychTools)

# Load plots
source("code/lecture_2_pca_plots.R")

style_mono_accent(
  # base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B05B", # DAPR3
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro"),
  text_font_size="1.3rem"
)
```

# Today's topics
+ Data reduction
+ Computing principal components
+ How many components to extract
+ Analytic rotation/transformation

---

# What's data reduction?
+ Mathematical and statistical procedures
  + Reduce large set of variables to a smaller set
  + Several forms of data reduction
	  + **Principal components analysis**
	  + **Factor analysis**
	  + Image analysis
	  + Correspondence analysis
	  + K-means clustering
	  + Multidimensional scaling
	  + Latent class analysis

---

# Data reduction versus other types of analyses
+ Like partial correlation, regression, ANOVA, and ANCOVA, you're making new variables by combining old ones


+ However, *no* variable is singled out as the "dependent variable"
  + In principal components analyses, there are no dependent variables
  + In factor analysis, *all* the variables are dependent variables

---

# When might you use data reduction?
+ You work with observational data and many variables
  + Psychology (differential, industrial/organizational)
  + Genetics
  + Epidemiology
  + Paleontology (yes, dinosaurs)

---

# Questions to ask before you start
+ Why are your variables correlated?
  + Agnostic/don't care
  + Believe there *are* underlying "causes" of these correlations 


+ What are your goals?
  + Just reduce the number of variables
  + Reduce your variables and learn about/model their underlying
  (latent) causes

---

# Questions to ask before you start
+ Why are your variables correlated?
  + **Agnostic/don't care**
  + Believe there *are* underlying "causes" of these correlations


+ What are your goals?
  + **Just reduce the number of variables**
  + Reduce your variables and learn about/model their underlying
  (latent) causes

---
 
# Principal components analysis
+ Goal is explaining as much of the total variance in a data set as possible
  + Starts with original data
  + Calculates covariances (correlations) between variables
  + Applies procedure called 'eigendecomposition' to calculate a set of linear composites of the original variables

---

<center>
# Caution!
</center>
+ Equations, math(s) even, ahead!
  + Point is *not* to memorize or be able to do these calculations
  + Point *is* to understand the background of these methods


- I will get back to this

---

# Linear composites
+ Linear composites are the principal components
  + If variables are closely related one another (large covariances/correlations), they can be represented by fewer composites, or even one composite
  + If variables are not closely related (small covariances/correlations), a single composite will not do a good job of representing them


+ To see why, we need to look at eigendecomposition

---

# Eigendecomposition
+ Our linear composite, or principal component, $\mathbf{z}$, is equal to: ${x_1}{w_1} + \dots + {x_p}{w_p}$
	+ $x_p$ are the variables and $w_p$ are weights that relate the variables to the component


+ We can express this using matrix algebra notation, $\mathbf{z} = \mathbf{X}\mathbf{w}$
	+ $\mathbf{X}$ is a data matrix with $p$ columns (one per
	variable) and $n$ rows (one per participant)
	+ $\mathbf{w}$ is the *eigenvector*
		+ Single column (vector) that contains the weights
          $(w_1,\dots, w_p)'$
		+ Maximize the variance explained by $\mathbf{z}$


+ The multiple regression equivalents of these equations are:
$$\mathbf{\hat{y}} = {x_1}{\beta_1} + \dots + {x_p}{\beta_p}$$
$$\mathbf{\hat{y}} = \mathbf{X}\mathbf{\beta}$$

---

# Eigendecomposition
+ To figure out $\mathbf{z}$, we need to solve for $\mathbf{w}$


+ For this purpose, we exploit a well-known relationship,
  $\mathbf{A}\mathbf{w} = \lambda\mathbf{w}$
	  + $\mathbf{A}$ is a $p \times p$ matrix (the correlation or
      covariance matrix from your data)
	  + $\lambda$ is the *eigenvalue* for the eigenvector $\mathbf{w}$


+ What this means is that...
	+ a vector of a square matrix is an eigenvector if it has a
    corresponding eigenvalue
	+ a vector is an eigenvector if, when we multiply it by a square
      matrix, that vector has been scaled
	+ the eigenvalue is the variance of the component


+ A $p \times p$ matrix has $p$ eigenvectors, each of which has $p$ elements,
  and $p$ eigenvalues, one for each eigenvector

---

```{r Eigenpicture, echo=FALSE, fig.align="center"}
plot(c(0,1), c(0,1), xlim=c(0,2), ylim=c(0,2),
     xlab="x", ylab="y")

arrows(0,0,1,1, col="blue", lwd=3)

arrows(1,1,2,2, col="red", lwd=3)

```

<center> An eigenvalue of 2 doubles the length of the corresponding
eigenvector (the blue arrow) </center>

---

# Solving for eigenvectors
+ Start by rearranging the equation, so that it is $(\mathbf{A}-\lambda\mathbf{I})\mathbf{w}=0$
  + The math is involved, but you can use R's `eigen` function to
    solve it
  + Consider two variables, x1 and x2, with an $r=0.80$
  
```{r Eigenvalues and vectors}
# Create the 2 x 2 matrix that shows the correlation of .8
mat <- matrix(c(1, .8, .8, 1), nrow=2)

# Use the eigen function to get the 2 eigenvectors and 2 eigenvalues
eigen(mat)
```
---

# What PCA with two variables, *r* = 0.00, looks like
The eigenvectors and eigenvalues are:

```{r eigendecomposition}
# Because correlation is 0, I had to add a tiny constant. See what
# happens if you don't do this.
mat <- matrix(c(1, -0.000001, -0.000001, 1), nrow=2)

# Use the eigen function to get the 2 eigenvectors and 2 eigenvalues
eigen(mat)
```
---

# Plot raw data
```{r echo=FALSE, out.height="450", out.width="450", fig.align="center"}
plot(g1vec)
```

---

# Add both (*p* = 2) eigenvectors
```{r echo=FALSE, out.height="450", out.width="450", fig.align="center"}
plot(g1vec2)
```

---

# What eigenvalues do when *r* = 0.00
```{r echo=FALSE, out.height="450", out.width="450", fig.align="center"}
plot(g1val2)
```

---

# Eigenvector 1 maximizes variance explained
```{r echo=FALSE, out.height="450", out.width="450", fig.align="center"}
plot(g1val3)
```

---

# Eigenvector 2 maximizes variance explained
```{r echo=FALSE, out.height="450", out.width="450", fig.align="center"}
plot(g1val4)
```
---

# What PCA with two variables, *r* = 0.80, looks like
+ Remember how we obtained this before?

```{r Eigenvalues and vectors 2}
# The 2 x 2 matrix that shows the correlation of .8
mat <- matrix(c(1, .8, .8, 1), nrow=2)

# Applying the eigen function to the correlation matrix
eigen(mat)

```

---

# Plot raw data
```{r echo=FALSE, out.height="450", out.width="450", fig.align="center"}
plot(g2vec)
```

---

# Add eigenvectors
```{r echo=FALSE, out.height="450", out.width="450", fig.align="center"}
plot(g2vec2)
```

---

# What eigenvalues do when *r* = 0.80
```{r echo=FALSE, out.height="450", out.width="450", fig.align="center"}
plot(g2val2)
```

---

# Eigenvector 1 maximizes variance explained
```{r echo=FALSE, out.height="450", out.width="450", fig.align="center"}
plot(g2val3)
```

---

# Eigenvector 2 maximizes variance explained
```{r echo=FALSE, out.height="450", out.width="450", fig.align="center"}
plot(g2val4)
```
---

# Eigenvalues and variance
+ In each situation, the eigenvalues we obtained for the correlation
  ended up being equal to the number of variables
  + $r=0.00$: $\lambda_{1}+\lambda_{2}=1+1=2=p$
  + $r=0.80$: $\lambda_{1}+\lambda_{2}=1.8+0.2=2=p$


+ This is a general rule for eigenvalues
	+ It has to do with the matrix algebra of eigendecomposition
	+ You don't need to know the details, but they can be made
      available if you're interested
---

# Eigenvalues and variances
+ Eigendecomposition has transformed the data, but the variance is the same
  + Total variance is the sum of variances of your variables
  + Sum of eigenvalues therefore must equal the total variance
  + Another way to state this is:

$$\text{total variance} =
\Sigma\lambda_{1}\lambda_{2}\cdots\lambda_{p}=p$$


+ Any component $m$ therefore accounts for a proportion of variance equal to:
$$\frac{\lambda_{m}}{\Sigma\lambda_{1}\lambda_{2}\cdots\lambda_{p}}=\frac{\text{component }m\text{'s eigenvalue}}{\text{total variance}}=\frac{\text{component }m\text{'s eigenvalue}}{p}$$

---

# Eigenvalues and variances
+ Let's apply these formula to the two principal components analyses
  we conducted
  + $r=0.00$
	  + Component 1 accounts for $1.00/2.00=0.50$ or 50% of the variance
	  + Component 2 accounts for $1.00/2.00=0.50$ or 50% of the variance
	+ $r=0.80$
	  + Component 1 accounts for $1.80/2.00=0.90$ or 90% of the variance
	  + Component 2 accounts for $0.20/2.00=0.10$ or 10% of the variance
		

+ Bottom line: When variables are correlated, a smaller subset of
  components can account for most of the variance, hence why principal
  components analysis is called *data reduction*

---

# Features of principal components analysis solutions
+ There are many components as original variables, i.e., $p$
  + This doesn't seem to make sense; wasn't the goal data reduction?
  + Makes sense when you consider that the goal of PCA was to explain
	*all* of the variance; doing that requires "extracting" $p$
	components
+ Each component maximizes the amount of variance explained
  + Components therefore decrease with respect to explained
  variance
  + Component 1 accounts for more variance than component 2, which
  explains more than component 3, etc. What do you think the pattern
  is with eigenvalues? 
+ Component are uncorrelated with one another
  + You saw this in the plots: the axes were at right angles, that is,
  *orthogonal*
  + An $r=0.00$ can be represented in geometry by a 90-degree angle

---
# Obtaining linear composites
+ A lot like calculating $\hat{y}$ (predicted values) in linear
  regression
  + Weights are calculated, multiplied by observed values on the
  variables for an individual, and summed
  + Every individual now has a score on the component(s)
  + These weights were in our eigenvector, so the equation for
  calculating everybody's score on the first component is:

$$\mathbf{z}_{1}={x_1}{w_{11}}+{x_1}{w_{21}}+\dots+{w_{p1}}{x_p}$$
	
---
# Obtaining *standardized* linear composites
+ The previous formulation gives you raw scores
  + Earlier component scores will have more variance than later
    component scores
  + For various reasons, that's not desirable
  

+ We can get standardized scores by first standardizing our weights:
$$\tilde{w}_{ij}=\frac{w_{ij}}{\sqrt{\lambda_j}}$$ here, $w_{ij}$ is
an element in an eigenvector that relates variable $i$ to component
$j$

---

# Obtaining component *loadings*
+ Allow you to understand the structure of your questionnaire
  + Loadings are presented in a table of p rows and m columns
	  + Each cell indicates the strength of relationship between a
	 variable and a component
	  + Loadings range from -1 to +1
	  + The higher the *absolute* loading, the stronger the association
   + The sum of squared loadings for any variable on all components equals 1
   + All variance in the variable is explained if we retain all components

---

# Obtaining component *loadings*
+ To get factor loadings, we scale the multiply the element of an
  eigenvector relating variable $i$ to component $j$ by the square
  root of component $j$'s eigenvalue
$$w^{*}_{ij}=w_{ij}{\sqrt{\lambda_j}}$$


+ This scales the weights so that the resulting loadings of components
onto variables are higher for components that have higher eigenvalues


+ Another way to state this is that you're giving more weight to
 relationships between variables and components when those components
 account for more variance

---

<center>
# Don't Panic!
</center>
+ Remember, you won't have to manually compute *any* of this to run an analysis
or to complete an assignment let alone memorize this!
	+ The `principal` function, which was written for *R*, does this
      stuff for you 
	+ It's good to know these details so that you can understand, or at least
	get the gist of, what is being done to your data


+ The remaining topics *won't* be covered in so much detail

---
# Number of components
+ Recall that the goal of PCA is to explain *all* the variance in a
  set of items, however, to do so, you need to extract $p$ components.
  + This defeats the purpose of data *reduction*; we want to extract
    fewer components
	  + Doing this makes sense if we don't sacrifice *too much* in the
	  way of explaining variance
	  + Taking our example, explaining nine-tenths of the variance
      with half the variables is a great trade-off; however, it's
      usually not so clear
	  
	  
+ There are tools that can *help* us make this decision
  + I said 'help' because there is no "correct" number of components
  + We also need to rely also on our judgement
	+ In light of what we know, does the set of components make sense?
	+ Do they tell us anything about our theory about what we're measuring?

---

# Number of components
+ Different tools tend to lead to over- and under-extraction
  + Under-extraction
	  + Retaining fewer than the optimal number
	  + Components are harder to interpret; smaller components are
      "squashed together"
  + Over-extraction
	  + Retaining more than the optimal number
	  + Components retained with only small number of primary loadings
	  + Will often get components related to item difficulty and not the
      constructs of interest


+ Some tools
  + Kaiser criterion
  + The scree plot
  + Velicer's Minimum Average Partial method
  + Parallel analysis
  
---
# Simulation studies
+ Before moving on, it's worth knowing that much of what we know about
  principal components analysis and factor analysis comes from
  simulation studies
  + Researchers create datasets that have known properties (number of
    components, how closely components are related to items, etc.)
  + Test how adequately certain methods recover these properties
	

---
# The Kaiser criterion
+ Because there are as many eigenvalues as there are variables, the
  mean eigenvalue will be 1. Therefore, retain any component, $m$, if
  $\lambda_{m}>1$
  + This method performs $very$ poorly
  + Simulation studies show that this method nearly always leads to
    over-extraction
  + Yet it's *still* the default in a lot of stats packages!


+ **Never** use it!

---

# Scree plots
+ Scree is the rubble at the bottom of mountains and the like. If you
  visit the Crags or Arthur's Seat, you will see it
  + Plot the eigenvalues of the components in descending order
  + The point at which there are "diminishing returns" and it makes
  little sense to take out more components is the point at which there
  is an 'elbow' in the plot
  
  
  + Studies of simulated data find that the performance of scree plots
    has been found to be inconsistent
	+ Sometimes it is correct, sometimes it leads to under-extraction,
      and other times it leads to over-extraction
	+ This is because there is an element of subjectivity to
      interpreting a scree plot

+ Not bad if used alongside other methods, but you shouldn't rely on it alone!
	
---

```{r BFI, fig.align="center", out.height="350", out.width="350", fig.cap="Scree plot of personality data: Where is the elbow?"}

library(psych)

## Selecting participants who have complete Big Five Inventory data
bfi <- bfi[complete.cases(bfi[1:25]),1:25]

## Producing scree plot
scree(bfi, factors=F, main="")

```

---

# Minimum Average Partial (MAP) technique
+ The idea is that you want to stop extracting components when by doing
  so you're no longer removing systematic variance
  + To do this, you first extract one component, then two, then three,
  etc., until you have done this $p$-1 times.
  + After each step, you compute the average of the partial
  correlations between all of the variables
  + The number of components that yields the lowest value (the
    minimum) is said to be the number of components
		+ Extract too few and there will be remaining systematic
		variance and MAP will be higher
		+ Extract too many and you will be "fitting error"; also leads
          to a higher MAP


+ Simulation studies show that this method tends to lead to under-extraction

---

```{r MAP}
library(psych)

## Selecting participants who have complete Big Five Inventory data
bfi <- bfi[complete.cases(bfi[1:25]),1:25]

## Getting MAP results using VSS function in R and saving to object
map_results <- VSS(bfi, plot=FALSE, method="pc", n=24)$map

## Using the paste and which.min functions to report the number of
## components that yields the lowest MAP
paste("MAP is lowest for", which.min(map_results), "components")

```
---

# Parallel analysis
+ The idea here is to conduct principal components analyses on random
data that has the same number of participants and the same number of
items
  + After doing this many times (I do it 1000 times), plot the
  distribution of the eigenvalues from these data sets
  + Then compare eigenvalues from your real data to the random
  ones, retaining only the components from real data if their
  eigenvalues are higher than the random ones


+ Simulation studies show that parallel analysis does a good job of
  recovering the right number of components from artificial data

---

```{r parallel analysis, fig.align="center"}

fa.parallel(bfi[complete.cases(bfi[1:25]),1:25], fa="pc", main="", quant=.95)

```

---

# Number of components
+ There are more tools to help identify the number of components
  + All have advantages and disadvantages
  + Best strategy is to use multiple methods that you believe are
  robust given your situation
	  + How are the data distributed?
	  + Is there likely to be a problem with under- or over-extraction?
	  + How reliable are the data?
	  + Do I expect any correlated methods?
	  
	  
+ Most important is how consistent the components are with what's
  known and with theory

---

# Checking the first draft of your questionnaire
+ When checking data obtained using the first draft, one thing to
  check is whether the variables you chose to measure some construct
  *do* measure that construct alone
  + An initial indication comes from the component loadings
	+ Does component A load on your measures of construct X?
	+ Does component A load on your measures of constructs other than
	X?
	+ Do components other than A load on your measures of construct X?
 

+ Your answers should be 'yes', 'no', and 'no'. 
  + If they are not, then there is a problem with one or more items or
  with the theory from which the construct(s) were derived
  + Depending on what you find, you'll need to revise or throw out
    items, revise or throw out your theory, or both

---

## Example
+ We'll use PCA to examine the Big Five Inventory, which was developed
  based on a well-established model of human personality
  + Because there's not enough 'room', I will stick to the
    extraversion and neuroticism items
  + I'll use the `principal` function to do the analysis


+ We'll also learn about the loading matrix---the lambda
  matrix---along the way

---
|    |   PC1 |   PC2 | $h^2$ | $u^2$ |  com |
|:---|------:|------:|------:|------:|-----:|
| E1 |  0.44 | -0.55 |  0.50 |  0.50 | 1.91 |
| E2 |  0.66 | -0.43 |  0.63 |  0.37 | 1.72 |
| E3 | -0.42 |  0.55 |  0.48 |  0.52 | 1.88 |
| E4 | -0.57 |  0.48 |  0.56 |  0.44 | 1.94 |
| E5 | -0.40 |  0.54 |  0.45 |  0.55 | 1.84 |
| N1 |  0.65 |  0.54 |  0.71 |  0.29 | 1.94 |
| N2 |  0.64 |  0.52 |  0.68 |  0.32 | 1.91 |
| N3 |  0.66 |  0.47 |  0.66 |  0.34 | 1.81 |
| N4 |  0.72 |  0.14 |  0.54 |  0.46 | 1.07 |
| N5 |  0.57 |  0.28 |  0.41 |  0.59 | 1.45 |

PC1 and PC2 = component loadings; $h^2$ = communalities, $u^2$ =
uniquenesses, com = complexities

---

# Definitions
+ Loadings are the size of associations between items and components
  + How 'high' they should be before one considers them 'salient' depends on the research
  + In personality research, depending on how reliable your measures
        are considered 'salient' if they are $\ge$ |.3| or |.4|
+ Communalities are sum of squared loadings across each row
  + They tell you proportion of item variance that your components account for
  + Uniquenesses equal $1-h^2$
  + Complexity indicates the extent to which an item loads on one
    component only; maximum is equal to the number of components
+ R provides extra information that I didn't display
  + Various 'fit indices'
  + Proportion of variance accounted for by each component
	  + How do you think these are calculated?
	  + Hint: It's similar to how the communalities are calculated

---
|    |       PC1 |       PC2 | $h^2$ | $u^2$ |  com |
|:---|----------:|----------:|------:|------:|-----:|
| E1 |  **0.44** | **-0.55** |  0.50 |  0.50 | 1.91 |
| E2 |  **0.66** | **-0.43** |  0.63 |  0.37 | 1.72 |
| E3 | **-0.42** |  **0.55** |  0.48 |  0.52 | 1.88 |
| E4 | **-0.57** |  **0.48** |  0.56 |  0.44 | 1.94 |
| E5 | **-0.40** |  **0.54** |  0.45 |  0.55 | 1.84 |
| N1 |  **0.65** |  **0.54** |  0.71 |  0.29 | 1.94 |
| N2 |  **0.64** |  **0.52** |  0.68 |  0.32 | 1.91 |
| N3 |  **0.66** |  **0.47** |  0.66 |  0.34 | 1.81 |
| N4 |  **0.72** |      0.14 |  0.54 |  0.46 | 1.07 |
| N5 |  **0.57** |      0.28 |  0.41 |  0.59 | 1.45 |

Many items have salient loadings (highlighed in red) of similar magnitudes on both
components. This is known as having 'cross loadings'. It's often not ideal.

---

# Analytic rotation
+ You'd be forgiven for thinking that the *cross-loadings* from the
  previous solution suggest it's time to revise/throw-out items or
  your theory
  + Don't panic!
  + Can happen because the initial "position" of the axes is arbitrary
  + Default of some statistics packages
	+ *Not* the default for the `principal` function
	+ I told R to keep the arbitrary position


+ If this is the problem, we can change the position of (rotate)
  the reference axes

---

|    |      RC1 |       RC2 | $h^2$ | $u^2$ |  com |
|:---|---------:|----------:|------:|------:|-----:|
| E1 |     0.01 | **-0.71** |  0.50 |  0.50 | 1.00 |
| E2 |     0.26 | **-0.75** |  0.63 |  0.37 | 1.24 |
| E3 |     0.00 |  **0.69** |  0.48 |  0.52 | 1.00 |
| E4 |    -0.16 |  **0.73** |  0.56 |  0.44 | 1.09 |
| E5 |     0.01 |  **0.67** |  0.45 |  0.55 | 1.00 |
| N1 | **0.84** |      0.03 |  0.71 |  0.29 | 1.00 |
| N2 | **0.82** |      0.02 |  0.68 |  0.32 | 1.00 |
| N3 | **0.81** |     -0.03 |  0.66 |  0.34 | 1.00 |
| N4 | **0.66** |     -0.33 |  0.54 |  0.46 | 1.48 |
| N5 | **0.63** |     -0.13 |  0.41 |  0.59 | 1.09 |

There are next to no cross-loadings and complexities are overall
low; N4: 'Often feel blue' is a possible exception, although it *is*
consistent with the association between extraversion and higher well-being

---

# Analytic rotation
+ Categories
  + Orthogonal
	  + Includes varimax and quartimax rotations
	  + Axes at right angles; correlations between components are zero
	  + In example, I used a varimax rotation

---

```{r rotation_2, echo=FALSE, fig.align="center"}

unrotated <- principal(bfi[complete.cases(bfi[1:25]),11:20],
                       nfactors=2, rotate="none")

rotated <- principal(bfi[complete.cases(bfi[1:25]),11:20],
                       nfactors=2, rotate="varimax")

## Plot factor loadings and unrotated axes
plot(unrotated$loadings, pch=19, col="black", cex=4, main="Varimax rotation")

text(unrotated$loadings, labels=rownames(unrotated$loadings), col="white")

abline(a=0, b=0, col="red")

abline(v=0, col="green")

## Varimax: Obtain angle between original axis 1 and new axis 1
angle_1old_1new <- 180*acos(rotated$rot.mat[1,1])/pi

## Varimax: Obtain angle between original axis 1 and new axis 2
angle_1old_2new <- 180*acos(rotated$rot.mat[1,2])/pi

## Add rotated axes to plot
abline(a=0, b=tan(angle_1old_1new*pi/180), col="red", lty="dashed")

abline(a=0, b=tan(angle_1old_2new*pi/180), col="green", lty="dashed")

legend("bottomleft",
  legend = c("Unrotated", "Rotated"),
  col = c("black", "black"),
  lty = c("solid", "dashed"),
  bty = "n")

```
<center>
What could you do to bring axes even closer to the points?
</center>

---

# Analytic rotation
+ Categories
  + Orthogonal
	  + Includes varimax and quartimax rotations
	  + Axes at right angles; correlations between components are zero
	  + In example, I used a varimax rotation
   + Oblique
	 + Includes promax and oblimin rotations
	 + Axes are not at right angles; correlations between components
       are not zero
	 + What happens when we apply an oblimin "rotation"?

---

|    |      TC1 |       TC2 | $h^2$ | $u^2$ | com |
|:---|---------:|----------:|------:|------:|----:|
| E1 |    -0.08 | **-0.72** |  0.50 |  0.50 | 1.0 |
| E2 |     0.16 | **-0.75** |  0.63 |  0.37 | 1.1 |
| E3 |     0.09 |  **0.70** |  0.48 |  0.52 | 1.0 |
| E4 |    -0.06 |  **0.74** |  0.56 |  0.44 | 1.0 |
| E5 |     0.10 |  **0.68** |  0.45 |  0.55 | 1.0 |
| N1 | **0.85** |      0.07 |  0.71 |  0.29 | 1.0 |
| N2 | **0.83** |      0.06 |  0.68 |  0.32 | 1.0 |
| N3 | **0.82** |      0.01 |  0.66 |  0.34 | 1.0 |
| N4 | **0.62** |     -0.31 |  0.54 |  0.46 | 1.5 |
| N5 | **0.61** |     -0.10 |  0.41 |  0.59 | 1.1 |

Again, there are next to no cross-loadings and complexities are
overall low; compared to the varimax structure, in most cases, the
high loadings are closer to 1.00 and the low loadings are closer to
0.00

---
```{r rotation_3, echo=FALSE, fig.align="center"}

## Plot factor loadings and unrotated axes
plot(unrotated$loadings, pch=19, col="black", cex=4,
     main="Varimax and oblimin rotations")

text(unrotated$loadings, labels=rownames(unrotated$loadings), col="white")

## Get oblimin-rotated components
rotated_ob <- principal(bfi[complete.cases(bfi[1:25]),11:20],
                       nfactors=2, rotate="oblimin")

## Oblimin: Obtain angle between original axis 1 and new axis 1
angle_1old_1new_ob <- 180*acos(rotated_ob$rot.mat[1,1])/pi

## Oblimin: Obtain angle between original axis 1 and new axis 2
angle_1old_2new_ob <- 180*acos(rotated_ob$rot.mat[1,2])/pi

## Add axes to plot
abline(a=0, b=tan(angle_1old_1new*pi/180), col="red", lty="dashed")

abline(a=0, b=tan(angle_1old_2new*pi/180), col="green", lty="dashed")

abline(a=0, b=tan(angle_1old_1new_ob*pi/180), col="red", lty="dotted")

abline(a=0, b=tan(angle_1old_2new_ob*pi/180), col="green", lty="dotted")

legend("bottomleft",
  legend = c("Varimax", "Oblimin"),
  col = c("black", "black"),
  lty = c("dashed", "dotted"),
  bty = "n")

```

---
# Analytic rotation
+ Rotation is about finding the most interpretable set of components


+ There is not a "unique" solution to this problem
  + Rotations are mathematically indistinguishable 
  + All you're doing is changing the reference point
  + This problem is known as "rotational indeterminacy"
  
  
+ "Simple structure" was propopsed by Thurstone as the criterion
  + Different means of rotation try to achieve this differently
  + I won't get into the maths!

---

# Simple structure
+ Adapted from Sass and Schmitt 2011

	1. Each variable (row) should have at least one zero loading 

	2. Each component (column) should have same number of zero’s as there are
   components
   
   3. Every pair of components (columns) should have several variables which
   load on one component, but not the other 
   
   4. Whenever more than four components are extracted, each pair of
   components (columns) should have a large proportion of variables
   which do not load on either component
   
   5. Every pair of components should have few variables which load on both
   components
  
---

# Question to ponder (and post about)
+ After rotation, the `principal` function labels each component a 
  rotated component (RC) instead of a principal component (PC)
  + This is because after rotating principal components, they are no
  longer principal components?
  + Hint: Think about what a principal component tries to do and then
    see whether the rotated components manage to do that
---

# Summary
+ You can use eigendecomposition to compute principal components,
  which can summarize your data using fewer variables and without
  sacrificing too much explanatory power
  + Rotating components is often necessary to clarify their meaning
  + Your judgement is involved in *all* major steps


+ Principal components analysis has many benefits
  + By using these components instead of your raw data, you reduce
  the Type I error rate
  + You can find poorly-behaving items and learn something about what
    it is you're measuring
