---
title: "<b>Week 7: Experimental Design Recap </b>"
subtitle: "RMS2<br><br> "
author: ""
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
```


# Topics for this video

+ Recap the key features and terminology of experimental design and ANOVA:
  + One-way & factorial designs
  + Factors, levels etc.
  + Between & within person
  + Covariates (ANCOVA)
  
+ If all of the above are familiar to you, you can skip this video.

---
# Experimental Design: manipulation
+ A key feature of experimental designs is that we actively manipulate our predictor (IV).

+ The intention is that changing the predictor will result in changes in the outcome (DV).

+ That is our manipulation will lead to variation in the outcome.

+ Our experiments can fail because we design these manipulations poorly.

+ The predictors in an experiment are (primarily) experimental conditions.

---
# Conditions/Factors & levels
+ **Conditions**: 
  + Are part of our experimental designs.
  + They are what is manipulated.
  
+ **Factors**
  + The resultant variables in our data set that code the experimental conditions are typically called factors.
  + Sometimes these are used interchangeably.
  + But it is useful to differentiate the design (conditions) and the data (factors)

+ Factors can have **levels**
  + These are the number of ways we vary or manipulate our condition

---
# Between vs Within Person
+ Two broad choices of study structure:
  + **Between person**: Participants only appear on one level/condition
  + **Within person**: Participants appear in multiple level/conditions

--

+ Study design labels reflect the between vs within and the number of conditions.
  + One-way between person
  + Two-way within person
  + etc.

---
# A new study
+ Suppose we wanted to look at the number of reading errors caused by noise distraction. 

+ We might devise a task where participants had to read a passage of text and put a cross through all verbs. 

+ Our outcome, or dependent variable, is the number of verbs correctly crossed out.

---
# One-way Between Person

```{css, echo=F}
    .remark-slide thead, .remark-slide tr:nth-child(2n) {
        background-color: white;
    }
```

```{r, echo=FALSE}
tibble(
  None = c("Tom", "Aja", "Alex", "Brandy", "Darren", "Lucy"),
  Noise = c("Adam", "Fiona", "Simon", "Tasha", "Josh", "Charlotte")
) %>%
  kable(.) %>%
  kable_styling(full_width = F) %>%
  add_header_above(c("Noise Level" = 2))
```


???
0db
85db

---
# One-way Between Person (more levels)


```{r, echo=FALSE}
tibble(
  None = c("Tom", "Aja", "Alex", "Brandy"),
  Moderate = c("Darren", "Lucy", "Josh", "Charlotte"),
  Loud = c("Adam", "Fiona", "Simon", "Tasha")
) %>%
  kable(.) %>%
  kable_styling(full_width = F) %>%
  add_header_above(c("Noise Level" = 3))
```

???
0db
85db
100db

---
# Two-way Between Person

```{r, echo=FALSE}
tibble(
  Distraction = c("Words", "Words", "No Words", "No Words"),
  None = c("Tom", "Aja", "Alex", "Brandy"),
  Moderate = c("Darren", "Lucy", "Josh", "Charlotte"),
  Loud = c("Adam", "Fiona", "Simon", "Tasha")
) %>%
  kable(.) %>%
  kable_styling(full_width = F) %>%
  row_spec(1:2, background = "lightblue") %>%
  add_header_above(c("", "Noise" = 3)) %>%
  collapse_rows(1, valign = "top")
```

???
0db
85db
100db

As we have a verbal task, possible noise with words would be a greater distraction



---
# One-way Within Person

```{r, echo=FALSE}
tibble(
  None = c("Tom", "Aja", "Alex", "Brandy", "Darren", "Lucy",  "Josh", "Charlotte"),
  Noise = c("Tom", "Aja", "Alex", "Brandy", "Darren", "Lucy",  "Josh", "Charlotte")
) %>%
  kable(.) %>%
  kable_styling(full_width = F) %>%
  add_header_above(c("Noise Level" = 2))
```




---
# Two-way Within Person

```{r, echo=FALSE}
tibble(
  Distraction = c(rep("Word", 4), rep("No Word", 4)),
  None = c(rep(c("Tom", "Aja", "Alex", "Brandy"),2)),
  Noise = c(rep(c("Tom", "Aja", "Alex", "Brandy"),2))
) %>%
  kable(.) %>%
  kable_styling(full_width = F) %>%
  row_spec(1:4, background = "lightblue") %>%
  add_header_above(c("", "Noise Level" = 2)) %>%
  collapse_rows(1, valign = "top")
```


---
# Mixed Designs
```{r, echo=FALSE}
tibble(
  Distraction = c(rep("Word", 4), rep("No Word", 4)),
  None = c("Tom", "Aja", "Alex", "Brandy", "Darren", "Lucy",  "Josh", "Charlotte"),
  Noise = c("Tom", "Aja", "Alex", "Brandy", "Darren", "Lucy",  "Josh", "Charlotte")
) %>%
  kable(.) %>%
  kable_styling(full_width = F) %>%
  row_spec(1:4, background = "lightblue") %>%
  add_header_above(c("", "Noise Level" = 2)) %>%
  collapse_rows(1, valign = "top")
```


---
# ANCOVA
+ When we conduct an experiment, we hope that via the processes of randomization of participants into conditions, we control for covariates or nuisance factors.

+ There are various types of design (e.g. randomized block designs) where explicit consideration of covariates is made during group assignment.

+ If we do not have key information before hand, we can measure covariates and include them.
  + Consider age in our noise example.
  
+ The resulting analysis model is called ANCOVA (within the ANOVA tradition)

---
# Summary 

+ We recapped key language for experimental design and ANOVA

+ Key points:
  + Experimental designs yield nominal categorical variables
  + These variables code groups.