---
title: "<b>Course Introduction </b>"
subtitle: "RMS2<br><br> "
author: "STATS TEAM"
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #  base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)

library(tidyverse)
```


```{r welcome, echo=FALSE}
cohort <- 2019
team <- c("Tom", "Anastasia", "Emma" , "Esperanza", 
          "Carolane", "Catherine", "Abby", "Svenja",
          "Yangzi", "Diana")
lecture_hours <- 10+10+50
lab_hours <- 20+20
```

# In this intro...
+ The RMS2 team

--

+ What each week will look like.

--

+ Expectations


---
# The Team
+ Tom Booth (CO)
+ Alex Doumas
+ Alex Weiss
+ Jo King
+ Umberto Noe
+ Aja Murray
+ Wonderful tutors (lots)

---
# RMS2 
+ In RMS2, we will teach you how to...

--

  + Run and interpret linear models for observational and experimental designs

--

  + Run and interpret repeated measures designs
  
--

  + Run and interpret factor analysis and principal components analysis


---
# Learning Outcomes
1. Understand multiple regression for continuous outcomes including interactions, interpretation, model assumptions, and when models should be applied.

2. Understand the relationship between ANOVA and regression under the general linear model and be able to implement basic coding schemes for categorical predictors.

3. Understand the principles of scale construction and data reduction methods including interpretation, model assumptions and methods to assess them.

4. Use R to practically conduct the analytic methods taught in the course.

5. Present and interpret the results of the analytic methods taught in the course.


---
# Who you are...

+ Psychology Y3 students

--

+ Psychology Y4 students who were abroad in 19-20

--

+ Intercalating medics

--

+ Perhaps some visitors from overseas

--

+ And there are a lot of you!



---
# The Structure
+ We are on-line, so this will look a little different to normal.

--

+ **Lectures**
  + Equivalent of ~2hr of live lecture, broken into shorter chunks
  + The chunks will be linked with tasks.
  + Some weeks additional background and recap material will also be provided.

+ **Independent study labs**
  + Structured reading and exercises with R
  + Equivalent of ~2-3 hrs per week
  
--

+ **Supported by**
  + In person on-line office hours (see LEARN for times)
  + Discussion boards

???
+ Requires you to work steadily and consistently across the year. 
+ It builds, and it is hard to cram.
+ To help you build this study skills, course has a specific assessment structure.


---
# The Assessment 
+ **Weekly quizzes** (25%)
  + 10 quizzes in total 
  + Quiz 1 is a practices.
  + The rest comprise your grade.
  + Mark is the average of your best 7/9 scores.

--

+ **Coursework report** (75%)
  + Provided a dataset and some questions
  + Task is to construct models to answer the questions, describe them, run them, report the results.
  + Will require use of multiple approaches from class

???
Key point to make is the little and often principle

---
# Materials
+ All of your materials for each week will be within the weekly folders on LEARN.

--

+ It will contain:
  + A playlist of recordings
  + Slide decks
  + Links to lab material
  + Links to the weekly quizzes
  + Any comments or specific instructions for the week
  + Links to discussion boards
  + Reading (all freely available online)

--

+ The weekly folders will become available incrementally.

---
# What you can expect from us
1. We will work hard to help you learn.

2. We will be open and communicate with you.

3. We will be polite, respectful and treat you like adults.


---
# What we expect of you
1. You work hard.

2. That you talk to me and the teaching team.

3. That you are polite, and respect the teaching team and your classmates.

4. Try and have fun.


---
# A brief word on engagement
+ It is going to be a different year for everyone

--

+ We are very keen to make sure everyone succeeds. 

--

+ To do that, we need to know when people are struggling. 

--

+ We will keep an eye on this in a number of ways:
  + Presence in OH
  + Engagement with material (watching recordings, discussion boards etc.)
  + Weekly quiz performance.
  
---
class: center, middle
# And that is it for this intro!
