---
title: "<b>Quick Quiz</b>"
subtitle: "<br><br> "
author: ""
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```


# Instructions
+ There will now be a small set of questions. 

--

+ Pause the video to think about them and write your answers down. 

--

+ The answers will be shown immediately afterwards.

--

+ This is to check learning and is not assessed.

---
# Question 1

+ Are these models nested?

```{r, eval=FALSE}

m0 <- lm(health ~ Conscientiousness + Neuroticism, data = data)

m1 <- lm(health ~ Conscientiousness + Extraversion, data = data)

```



---
# Question 1: Answer
+ Are these models nested?

```{r, eval=FALSE}

m0 <- lm(health ~ Conscientiousness + Neuroticism, data = data)

m1 <- lm(health ~ Conscientiousness + Extraversion, data = data)

```

+ No.
  + Like the lecture example, both models have unique variables
  + `m0` = Neuroticism
  + `m1` = Extraversion

---
# Question 2
+ Are these models nested?

```{r, eval=FALSE}

m0 <- lm(health ~ Age + Conscientiousness + Neuroticism, data = data)

m1 <- lm(health ~ Age + Conscientiousness + Neuroticism + Agreeableness, data = data)

```


---
# Question 2: Answer
+ Are these models nested?

```{r, eval=FALSE}

m0 <- lm(health ~ Age + Conscientiousness + Neuroticism, data = data)

m1 <- lm(health ~ Age + Conscientiousness + Neuroticism + Agreeableness, data = data)

```

+ Yes
  + The first 3 variables are the same in both models
  + `m1` adds agreeableness
  + Another way to think about this is that if we fixed the effect of Agreeableness to 0, `m0` = `m1`

---
# Question 3
+ Are these models nested?

```{r, eval=FALSE}

m0 <- lm(health ~ Age + Conscientiousness + Neuroticism, data = data)

m1 <- lm(mortality ~ Age + Conscientiousness + Neuroticism + Agreeableness, data = data)

```

---
# Question 3: Answer
+ Are these models nested?

```{r, eval=FALSE}

m0 <- lm(health ~ Age + Conscientiousness + Neuroticism, data = data)

m1 <- lm(mortality ~ Age + Conscientiousness + Neuroticism + Agreeableness, data = data)

```

+ No.
  + Did you say yes looking at Agreeableness being the only new predictor?
  + Look at the outcome...
  + Models can only be nested if the outcome is the same
  
---
# The end.

+ If you are happy with your answers, move on to the next video/task.

+ If you are unclear on anything, either pop along to office hours, or post a question on the discussion board. 