---
title: "<b>Week 8: ANOVA as lm - Assumptions </b>"
subtitle: "RMS2<BR><BR> "
author: ""
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
```


# Topics for today

+ A brief word about assumptions in linear models for factorial designs.

---
# Data and model

+ A little reminder of our model:

```{r}
hosp_tbl <- read_csv("hospital.csv", col_types = "dff")
m4 <- lm(SWB ~ Treatment + Hospital + Treatment*Hospital, data = hosp_tbl)
```

---
# Assumptions

+ And a reminder of linear model assumptions:
  + **L**inearity: The relationship between $y$ and $x$ is linear.
  + **I**ndependence of errors: The error terms should be independent from one another.
  + **N**ormality: The errors $\epsilon$ are normally distributed
  + **E**qual variances ("Homoscedasticity"): The scale of the variability of the errors  $\epsilon$ is constant at all values of $x$.  

--

**That is, exactly the same assumptions as would be the case for any other linear model**

---
# Assumptions: Equal variances
```{r}
plot(m4, which = 1)
```

---
# Assumptions: Normality of residuals
```{r}
plot(m4, which = 2)
```

---
# Diagnostics: Influence
```{r}
plot(m4, which = 4)
```


---
# Summary of today

+ Briefly recapped model assumption testing for linear models

+ Discussed the key advantage in this respect from approach experimental design from a linear model perspective (rather than traditional ANOVA) 
