---
title: "<b>Quick Quiz</b>"
subtitle: "<br><br> "
author: ""
institute: "Department of Psychology<br>The University of Edinburgh"
date: "AY 2020-2021"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  #base_color = "#0F4C81", # DAPR1
   base_color = "#BF1932", # DAPR2
  # base_color = "#88B04B", # DAPR3 
  # base_color = "#FCBB06", # USMR
  # base_color = "#a41ae4", # MSMR
  header_color = "#000000",
  header_font_google = google_font("Source Sans Pro"),
  header_font_weight = 400,
  text_font_google = google_font("Source Sans Pro", "400", "400i", "600", "600i"),
  code_font_google = google_font("Source Code Pro")
)
```


# Instructions
+ There will now be a small set of questions. 

--

+ Pause the video to think about them and write your answers down. 

--

+ The answers will be shown immediately afterwards.

--

+ This is to check learning and is not assessed.

---
# Question 1

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
midus <- read_csv("MIDUS2.csv")
midus2 <- midus %>%
  select(1:4, 31:42) %>%
  mutate(
    PIL = rowMeans(.[grep("PIL", names(.))],na.rm=T)
  ) %>%
  select(1:4, 12:17) %>%
  drop_na(.)
```

+ Which model fits best?

```{r}
m0 <- lm(health ~ age + sex + O + C + E + A + N, data = midus2)
m1 <- lm(health ~ age + sex + O + C + E + A + N + PIL, data = midus2)
```

.pull-left[
```{r}
AIC(m0, m1)
```
```{r}
BIC(m0, m1)
```
]

.pull-right[
```{r}
anova(m0, m1)
```
]


---
# Question 1: Answer
+ The models are nested.
+ Incremental F is significant
+ AIC for `m1` is smaller
+ BIC for `m1` is smaller and > 10
+ Model `m1` looks like the better model.

---
# Question 2

+ Which model fits best?

```{r}
m0 <- lm(health ~ age + E, data = midus2)
m1 <- lm(health ~ age + E + sex, data = midus2)
```

.pull-left[
```{r}
AIC(m0, m1)
```
```{r}
BIC(m0, m1)
```
]

.pull-right[
```{r}
anova(m0, m1)
```
]

---
# Question 2: Answer
+ The models are nested.
+ Incremental F is not significant
+ AIC for `m1` is smaller
+ BIC for `m0` is smaller.
+ We have contrasting information, but overall it looks like `m0` is the better model.

---
# Question 3

+ Which model fits best?

```{r}
m0 <- lm(health ~ age + O + C, data = midus2)
m1 <- lm(health ~ age + E + A + N, data = midus2)
```

.pull-left[
```{r}
AIC(m0, m1)
```
```{r}
BIC(m0, m1)
```
]

.pull-right[
```{r}
anova(m0, m1)
```
]

---
# Question 3: Answer
+ The models are non-nested.
+ Therefore you should have ignored the incremental F
+ AIC for `m1` is smaller
+ BIC for `m1` is smaller and > 10
+ Model `m1` looks like the better model.
  
---
# The end.

+ If you are happy with your answers, move on to the next video/task.

+ If you are unclear on anything, either pop along to office hours, or post a question on the discussion board. 